---
title: "Analysis Notebook"
author: "Kira Goff"
date: "2026-07-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-libraries}
library(tidyverse)
library(gggenes)
library(ggrepel)
library(gridExtra)
library(cowplot)
```

```{r data-set-up}
# Source your standalone R scripts
source("R/parse_neighbourhood.R")
source("R/extract_neighbourhood.R")
source("R/plot_neighbourhood.R")

# Define paths
gbff_path    <- "git-ignore/peri.polished.gbff"
de_csv_path  <- "git-ignore/annotated_operons_clusters_stats.csv"
out_dir      <- "git-ignore/operon-results" # this can also be defined inside of a graphing call
```

# Data structure for input

GBFF/GBK: 
This can handle --compliant files or default Bakta/Prokka output.

Differential expression file:

Does it matter which program I use to generate differential expression data?
- It's agnostic! Choose your favourite.

How does it know what my time points are? How many can I have?
- It looks for columns formatted [metric]_[timepoint]
- This should work for any reasonable nubmer of time points

How does it parse columns?
- For time points, you can provide numbers or hours, e.g.
  - [metric]_0
  - [metric]_0h
- For metrics, it has regex parsing
  - for log 2 fold change, it accepts
    - lfc
    - log2fc
    - fold
    - fc
  - for significance, it accepts
    - padj
    - p_adj
    - qval
    - pval
- Other columns it's looking for:
    - locus_tag (required)
    - gene (a short form gene name)
    - product (a longer descriptor for tables)
    - operon (if using the pre-calculated 'operon' an 'gene' modes)


```{r}
genomic_df <- parse_gbff(gbff_path = gbff_path, de_csv_path = de_csv_path)
```

# Example A: Query by predefined operons or locus tags

```{r query-by-operon}
 filtered_data <- parse_neighbourhood(
   master_data = genomic_df, 
   query_type = "operon",             # locus_tag for a specific gene
   query_values = c("2038"), 
   flank_operons = 2
 )
 
 plot_neighbourhood(
   filtered_data = filtered_data, 
   style = "above", 
   deseq_path = de_csv_path, 
   output_dir = out_dir,
 )
```

# Example B: Autodected operons 

This autodetects operons based on 
a) strand, and 
b) intergenic distance (default 100, but can be adjusted)

It's not as trustworthy as using something like operon-mapper, but it will suffice for visualization

```{r automated-operon-detection}
filtered_data <- parse_neighbourhood(
  master_data = genomic_df, 
  query_type = "operon_auto", 
  query_values = c("PBUTOS_03315"), # Gene of interest
  flank_operons = 0,               # flank operons upstream/downstream, if 0, autodetects gene operon
  max_gap = 100                    # Basepair gap limit to define an operon break, default = 100
)

plot_neighbourhood(
  filtered_data = filtered_data, 
  style = "none", 
  deseq_path = de_csv_path, 
  output_dir = out_dir,
  p_thresh = 0.1,           # significance threshold for * marker, default 0.05
  clean_short_genes = FALSE # combined with 'none' or 'table', sticks the names into the gene arrows
)
```

# Example C: Customizing visual outputs


```{r visual-customization}
filtered_data <- parse_neighbourhood(
  master_data = genomic_df, 
  query_type = "operon", 
  query_values = c("2922"), 
  flank_operons = 3,              
)

plot_neighbourhood(
  filtered_data = filtered_data, 
  style = "table", 
  deseq_path = de_csv_path, 
  output_dir = out_dir,
  width = 10,               # save width (in)
  height = 15,              # save height (in)
  colour_limits = c(-2,2),  # sets absolute colour limits for comparing multiple plots
  custom_palette = c("navy","lightgrey","darkorange3") # set low, mid, and high point gradients
)
```